---
title: "5. Spatial interpolation"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Spatial interpolation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette teaches you how to spatially interpolate stress fields and 
display the lateral patterns of stress anomalies.

```{r setup, message=FALSE}
library(tectonicr)
library(ggplot2) # load ggplot library
```

```{r load_data}
data("san_andreas")

data("cpm_models")
por <- cpm_models[["NNR-MORVEL56"]] |>
  equivalent_rotation("na", "pa")
```


## Interpolation
### Geographic coordinate system
Spatial interpolation of stress data is based on the aforementioned metrics:

```{r interpolation}
mean_SH <- stress2grid(san_andreas, gridsize = 1, R_range = seq(50, 350, 100))
```

The default settings apply quality and inverse distance weighting of the mean, 
as well as a 25% cut-off for the standard deviation. 

> The `tectonicr` algorithm is a modified version of the MATLAB algorithm 
'stress2grid' by Ziegler and Heidbach (2017) but it is faster and provides 
additional features.  See `help(stress2grid)` for more 
details and settings, such as different statistics for calculating the 
orientation average, adjustments of the weighting power of the distance, 
quality or method.

The data can now be visualized:  
```{r plot, warning=FALSE, message=FALSE}
trajectories <- eulerpole_loxodromes(x = por, n = 40, cw = FALSE)

ggplot(mean_SH) +
  geom_sf(data = trajectories, lty = 2) +
  geom_azimuth(data = san_andreas, aes(lon, lat, angle = azi), radius = .25, color = "grey30") +
  geom_azimuth(aes(lon, lat, angle = azi, alpha = sd, color = mdr), lwd = 1) +
  coord_sf(xlim = range(san_andreas$lon), ylim = range(san_andreas$lat)) +
  scale_alpha(name = "Standard deviation", range = c(1, .25)) +
  scale_color_viridis_c(
    "Wavelength\n(R-normalized mean distance)",
    limits = c(0, 1),
    breaks = seq(0, 1, .25)
  ) +
  facet_wrap(~R)
```

### PoR coordinate system

The interpolated direction of far apart data points will suffer from distortions
due to the underlying projection. In order to prevent such effects, the 
interpolation can be done in the PoR reference frame where the direction stays 
constant no matter the distance between the data points. Assuming that the 
stress field is sourced by the plate boundary force, the model-based 
interpolation allows more reliable results for areas close to plate boundaries.

```{r interpolation_PoR}
mean_SH_PoR <- PoR_stress2grid(san_andreas, PoR = por, gridsize = 1, R_range = seq(50, 350, 100))
```

```{r plot2, warning=FALSE, message=FALSE}
ggplot(mean_SH_PoR) +
  geom_sf(data = trajectories, lty = 2) +
  geom_azimuth(data = san_andreas, aes(lon, lat, angle = azi), radius = .25, color = "grey30") +
  geom_azimuth(aes(lon, lat, angle = azi, alpha = sd, color = mdr), radius = .5, lwd = 1) +
  coord_sf(xlim = range(san_andreas$lon), ylim = range(san_andreas$lat)) +
  scale_alpha(name = "Standard deviation", range = c(1, .25)) +
  scale_color_viridis_c(
    "Wavelength\n(R-normalized mean distance)",
    limits = c(0, 1),
    breaks = seq(0, 1, .25)
  ) +
  facet_wrap(~R)
```

## Rasterize the interpolation

The function `compact_grid()` selects only data with the minimum search radius 
from interpolated layers with different search radii. Since the interpolation 
was performed in the PoR CRS, the interpolated azimuths are additionally given 
in the transformed azimuths. This allows to easily calculate misfits from 
predicted directions:

```{r compact}
mean_SH_PoR_reduced <- mean_SH_PoR |>
  compact_grid() |>
  dplyr::mutate(cdist = circular_distance(azi.PoR, 135))
```

Using `circular_distance()` in the example above, we can display the spatial 
patterns of the misfits of the stress direction from the predicted direction of 
the plate boundary force. Since the interpolation was performed in the PoR CRS, 
the grid is not composed of equally spaced grid cells in the geographic CRS. To 
rasterize such grids, we can, e.g., use Voronoi cells from the `ggforce` 
package.

```{r voronoi, warning=FALSE, message=FALSE}
ggplot(mean_SH_PoR_reduced) +
  ggforce::geom_voronoi_tile(
    aes(lon, lat, fill = cdist),
    max.radius = .7, normalize = FALSE
  ) +
  scale_fill_viridis_c("Angular distance", limits = c(0, 1)) +
  geom_sf(data = trajectories, lty = 2) +
  geom_azimuth(
    aes(lon, lat, angle = azi, alpha = sd),
    radius = .25, lwd = .2, colour = "white"
  ) +
  scale_alpha("Standard deviation", range = c(1, .25)) +
  coord_sf(xlim = range(san_andreas$lon), ylim = range(san_andreas$lat))
```

The map highlights **stress anomalies** which show misfits to the direction of 
tested plate boundary force.


# References
Mardia, K. V., and Jupp, P. E. (Eds.). (1999). "Directional Statistics" 
Hoboken, NJ, USA: John Wiley & Sons, Inc. 
doi: 10.1002/9780470316979.
<!--doi: [10.1002/9780470316979](https://doi.org/10.1002/9780470316979).-->

Ziegler, Moritz O., and Oliver Heidbach. 2017. “Manual of the Matlab Script Stress2Grid” 
GFZ German Research Centre for Geosciences; World Stress Map Technical Report 17-02. 
doi: [10.5880/wsm.2017.002](https://doi.org/10.5880/wsm.2017.002). 
