---
title: "Circular statistics"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Circular statistics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette teaches you how to retrieve statistical parameters of
orientation data, for example the mean direction of stress datasets.

```{r setup, echo=TRUE, message=FALSE}
library(tectonicr)
library(ggplot2) # load ggplot library
```

## Orientation data types

Circular orientation data are two-dimensional orientation vectors that
are either **directional** or **axial**.

-   **Directional** is $2\pi$-periodical, the angle $\alpha$ is
    non-symmetrical on a circle. The modulus of the angle is $2\pi$
    (360°): $ \alpha = \alpha\, \text{mod}\, 2\pi$. Directional data
    usually involve processes where objects or particles were
    transported from one known place to another. Examples are wind
    direction, bird vanishing angles, plate motion, and fault slip
    directions.
-   **Axial** data is $\pi$-periodical, i.e. each direction is
    considered as equivalent to the opposite direction, so that the
    angles $\alpha$ and $\alpha + 180^\circ$ are equivalent. The modulus
    of the angle is $\pi$ (180°):
    $ \alpha = \alpha\, \text{mod}\, \pi$. Data usually involves
    transport of objects where the origin is unknown or doesn't matter
    as the transport could be extended infinitely. Examples are glacial
    striae, mineral alignments (long-axis of grain or crystal shapes),
    and principal stress/strain axes (e.g. shortening or compression
    direction, including the angle of the maximum horizontal stress
    $\sigma_\text{Hmax}$).

> In (almost) every function in the **tectonicr** library, the `axial` argument
> specifies whether your angles `x` are axial (`TRUE`) or directional
> (`FALSE`). Because this package is mainly designed for analyzing stress orientations,
> the default is `axial=TRUE`.

## Mean direction

In case of axial data, the calculation of the mean of, say, 
of 35$^{\circ}$ and 355$^{\circ}$ should be 15 instead of 195$^{\circ}$.
**tectonicr** provides the circular mean (`circular_mean()`) and the
quasi-median (`circular_median()`) as metrics to describe average
direction:

```{r mean, echo=TRUE}
data("san_andreas")
circular_mean(san_andreas$azi)
circular_median(san_andreas$azi)
```

> Again: if your angles are directional, you need to specify the `axial` argument. 

Note the different results: 

```{r mean_directional, echo=TRUE}
circular_mean(san_andreas$azi, axial = FALSE)
circular_median(san_andreas$azi, axial = FALSE)
```

## Quality weighted mean direction

Because the stress data is heteroscedastic, the data with less precise
direction should have less impact on the final mean direction The
weighted mean or quasi-median uses the reported measurements linear
weighted by the inverse of the uncertainties:

```{r weighted, echo=TRUE}
w <- weighting(san_andreas$unc)

circular_mean(san_andreas$azi, w)
circular_median(san_andreas$azi, w)
```

The spread of directional data can be expressed by the standard
deviation (for the mean) or the quasi-interquartile range (for the
median):

```{r weighted_spread, echo=TRUE}
circular_sd(san_andreas$azi, w) # standard deviation
circular_IQR(san_andreas$azi, w) # interquartile range
```

## Statistics in the Pole of Rotation (PoR) reference frame

**NOTE:** Because the $\sigma_{SHmax}$ orientations are subjected to
angular distortions in the geographical coordinate system, it is
recommended to express statistical parameters using the transformed
orientations of the PoR reference frame.

```{r por, echo=TRUE}
data("cpm_models")
por <- cpm_models[["NNR-MORVEL56"]] |>
  equivalent_rotation("na", "pa")
san_andreas.por <- PoR_shmax(san_andreas, por, type = "right")
```

```{r por_stats, echo=TRUE}
circular_mean(san_andreas.por$azi.PoR, w)
circular_sd(san_andreas.por$azi.PoR, w)

circular_median(san_andreas.por$azi.PoR, w)
circular_IQR(san_andreas.por$azi.PoR, w)
```

The collected summary statistics can be quickly obtained by
`circular_summary()`:

```{r summary_stats, echo=TRUE}
circular_summary(san_andreas.por$azi.PoR, w, mode = TRUE)
```

The summary statistics additionally include the circular
quasi-quantiles, the variance, the skewness, the kurtosis, the mode, the
95% confidence angle, and the mean resultant length (R).

## Rose diagram

**tectonicr** provides a rose diagram, i.e. histogram for angular data.

```{r rose1, echo=TRUE}
rose(san_andreas$azi,
  weights = w, main = "North pole",
  dots = TRUE, stack = TRUE, dot_cex = 0.5, dot_pch = 21
)

# add the density curve
plot_density(san_andreas$azi, kappa = 20, col = "#51127CFF", shrink = 1.5)
```

The diagram shows the uncertainty-weighted frequencies (equal area rose
fans), the von Mises density distribution (blue curve), and the circular
mean (red line) incl. its 95% confidence interval (transparent red).

Showing the distribution of the transformed data gives the better
representation of the angle distribution as there is no angle distortion
due to the arbitrarily chosen geographic coordinate system.

```{r rose2, echo=TRUE}
rose(san_andreas.por$azi,
  weights = w, main = "PoR",
  dots = TRUE, stack = TRUE, dot_cex = 0.5, dot_pch = 21
)
plot_density(san_andreas.por$azi, kappa = 20, col = "#51127CFF", shrink = 1.5)

# show the predicted direction
rose_line(135, radius = 1.1, col = "#FB8861FF")
```

The green line shows the predicted direction.

## QQ Plot

The (linearised) circular QQ-Plot (`circular_qqplot()`) can be used to
visually assess whether our stress sample is drawn from an uniform
distribution or has a preferred orientation.

```{r qqplot, echo=TRUE}
circular_qqplot(san_andreas.por$azi.PoR)
```

Our data clearly deviates from the diagonal line, indicating the data is
not randomly distributed and has a strong preferred orientation around
the 50% quantile.

## Statistical tests

### Test for random distribution

Uniformly distributed orientation can be described by the *von Mises
distribution* (Mardia and Jupp, 1999). If the directions are distributed
randomly can be tested with the **Rayleigh Test**:

```{r random, echo=TRUE}
rayleigh_test(san_andreas.por$azi.PoR)
```

Here, the test rejects the Null Hypothesis (`statistic > p.value`). Thus
the $\sigma_{SHmax}$ directions have a preferred orientation.

Alternative statistical tests for circular uniformity are
`kuiper_test()` and `watson_test()`. Read `help()` for more details...

## Test for goodness-of-fit

Assuming a von Mises Distribution (circular normal distribution) of the
orientation data, a $100(1-\alpha)\%$ **confidence interval** can be
calculated (Mardia and Jupp, 1999):

```{r confidence, echo=TRUE}
confidence_interval(san_andreas.por$azi.PoR, conf.level = 0.95, w = w)
```

The prediction for the $\sigma_{SHmax}$ orientation is $135^{\circ}$.
Since the prediction lies within the confidence interval, it can be
concluded with 95% confidence that the orientations follow the predicted
trend of $\sigma_{SHmax}$.

The (weighted) **circular dispersion** of the orientation angles around
the prediction is another way of assessing the significance of a normal
distribution around a specified direction. It can be measured by:

```{r dispersion, echo=TRUE}
circular_dispersion(san_andreas.por$azi.PoR, y = 135, w = w)
```

The value of the dispersion ranges between 0 and 2.

The standard error and the confidence interval of the calculated
circular dispersion can be estimated by bootstrapping via:

```{r dispersion_MLE, echo=TRUE}
circular_dispersion_boot(san_andreas.por$azi.PoR, y = 135, w = w, R = 1000)
```

The statistical test for the goodness-of-fit is the (weighted)
**Rayleigh Test** with a specified mean direction (here the predicted
direction of $135^{\circ}$:

```{r rayleigh2, echo=TRUE}
weighted_rayleigh(san_andreas.por$azi.PoR, mu = 135, w = w)
```

Here, the Null Hypothesis is rejected, and thus, the alternative, i.e.
an non-uniform distribution with the predicted direction as the mean
cannot be excluded.

# References

Mardia, K. V., and Jupp, P. E. (Eds.). (1999). "Directional Statistics"
Hoboken, NJ, USA: John Wiley & Sons, Inc. doi: 10.1002/9780470316979.
<!--doi: [10.1002/9780470316979](https://doi.org/10.1002/9780470316979).-->

Ziegler, Moritz O., and Oliver Heidbach. 2017. “Manual of the Matlab
Script Stress2Grid” GFZ German Research Centre for Geosciences; World
Stress Map Technical Report 17-02. doi:
[10.5880/wsm.2017.002](https://doi.org/10.5880/wsm.2017.002).
