---
title: "Introduction"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`tectonicr` is a free and open-source R package for modeling and
analyzing the direction of the maximum horizontal stress based on the
empirical link between the direction of intraplate stress and the
direction of the relative motion of neighboring plates.

## Prerequisites

Before you start you need to have R and and R-compatible IDE installed
on your computer:

1.  You will need R installed on your computer. You can download R from
    the CRAN website: <https://cran.r-project.org/>

2.  Once you downloaded and successfully installed R, you can use R in
    your preferred IDE. I recommend installing and using
    [RStudio](https://posit.co/download/rstudio-desktop/). However,
    [Positron](https://positron.posit.co/) and
    [VisualStudio](https://visualstudio.microsoft.com/) are also great
    alternatives. You are also allowed to use R's native GUI or use R in
    your terminal only.

3.  Open your IDE and now you can install the {tectonicr} package by
    typing into the commando console the following code. This will
    install the package and every required library as well.

```{r installl, echo=FALSE, eval=FALSE}
install.packages("tectonicr")

# or install the current development version from github:
remotes::install_github('tobiste/tectonicr')
```

4.  Next you must "load" the library to use the functions of
    {tectonicr}:

```{r setup, warning=FALSE, message=FALSE}
library(tectonicr)
```

## Theoretical background

The theory of intraplate tectonics (Wdowinski 1998) allows for
calculating the first-order intraplate deformation induced by horizontal
displacement of deformable plate boundaries (Stephan et al., 2023). It
is based on empirical link between the directions of relative plate
motion and the displacement and deformation fields within a plate
interior adjacent to three types of deformable plate boundaries:
inward-, outward-, and tangential-displaced boundaries. The model
predicts the direction of intraplate displacement, displacement rate,
strain, and stress fields in terms of small circles, great circles, and
45$^{\circ}$ loxodromes around the pole of rotation of two adjacent
plates. According to the theory, the principal axis of the maximum
horizontal stress follows small circles for inward-displaced boundaries,
great circles for outward-displaced boundaries, and loxodromes for
tangential-displaced boundaries.

The theory assumes that the first-order intraplate deformation is
predominantly induced by horizontal forces acting on plate boundaries
and by buoyancy forces that arise from lateral density variations
between mid-ocean ridges and plate interiors (ridge push).

### Inward, Outward and Tangential Displaced Boundaries

**Inward-moving plate boundaries** induce compressional horizontal
tractions from the plate boundary towards the plate's interior along the
direction of relative plate motion. These compressional tractions are
produced by forces related to subduction, collision, or ridge-push.
Thus, stresses across convergent plate boundaries are characterized by
the dominance of thrusting or strike-slip faulting
($\sigma_1 \approx \sigma_{Hmax}$) with $\sigma_{Hmax}$ (maximum
horizontal stress) trending parallel to the plate convergence, i.e.
parallel to *small circles* around the pole of the relative plate motion
(pole of rotation, PoR).

**Outward moving plate boundaries** produce tensional tractions and
displacements directed away from the plate interior. Along spreading
ridges and intracontinental rifting stresses are dominated by normal
faulting ($\sigma_1 \approx \sigma_{vertical}$,
$\sigma_2 \approx \sigma_{Hmax}$) with $\sigma_{Hmax}$ trending
perpendicular to the plate motion trajectories (i.e. along *great
circles*). In the case of intracontinental setting, stresses and
displacements may be associated to slab-retreat, back-arc extension, or
the release of the excess of gravitational potential energy stored in
thickened crust through, e.g., gravitational collapse.

Along transform boundaries (**tangential displaced boundaries**), the
two neighboring plates exert shear tractions tangential to the
orientation of the boundary (Forsyth and Uyeda, 1975). Faulting and
displacement adjacent to these plate boundaries are characterized by
strike-slip parallel to the plate motion, and thus, the principal axes
of maximum and minimum stress are orientated at an angle of c.
45$^{\circ}$ to the plate motion trajectory. Geometrically,
$\sigma_{Hmax}$ direction follows along 45$^{\circ}$ *loxodromes* (lines
of constant bearing) which diverge ---depending on the sense of the
transform boundary--- clockwise or counterclockwise from the relative
PoR and intersect both small and great circles at an angle of
45$^{\circ}$.

## Theoretical direction of Horizontal Stress and Deviation From the Measured Stress

<!-- **tectonicr** basically calculates the bearing from a point of  -->

<!-- interest to a given Euler pole of the relative plate motion by the formula: -->

<!-- ```{=tex} -->

<!-- \begin{equation}  -->

<!--   \theta = \arctan2 (\sin \Delta\lambda * \cos\psi_2, \cos\psi_1 \sin\psi_1-\sin\psi_1 \cos\psi_2 \cos\Delta\lambda) -->

<!-- \end{equation} -->

<!-- ``` -->

<!-- where $\psi_1$, $\lambda_1$ are the longitude and latitude of the point,  -->

<!-- $\psi_2$, $\lambda_2$ the longitude and latitude of the Euler pole  -->

<!-- ($\Delta\lambda$ is the difference in longitude).  -->

<!-- > The bearing $\theta$ is the orientation of a great circle that passes through  -->

<!-- the point and the Euler pole. Thus, it reflects the theoretical direction of  -->

<!-- $\sigma_{Hmax}$ that follows great circles trajectories.  -->

<!-- Small circle directions then are $90^{\circ}$ and loxodrome directions  -->

```{=html}
<!-- are $\pm45^{\circ}$ to $\theta$. 

<!-- --- -->
```

Trajectories of theoretical directions can modeled by the following
steps:

First, we need to specify coordinates of the Pole of Rotation (PoR) to
get the directions of the great circles, small circles, and loxodromes
around the PoR at the given point (e.g. at 45$^{\circ}$N/20$^{\circ}$E).

For example, the PoR has the coordinates: 90$^{\circ}$N/0$^{\circ}$E.
Then $\sigma_{Hmax}$ following great and small circles and loxodromes
geometries can be modeled with `model_shmax()`:

```{r direction_of_plate_motion, echo=TRUE}
# Example:
point <- data.frame(lat = 45, lon = 20)
por <- data.frame(lat = 90, lon = 0)
model <- model_shmax(point, por)
print(model)
```

If there is an observed stress direction at the point, e.g. azimuth of
$\sigma_{Hmax}$ is 90$^{\circ}$, the deviation from the modeled stress
directions can be calculated through `deviation_shmax()`:

```{r deviation_of_plate_motion, echo=TRUE}
deviation <- deviation_shmax(model, 90)
print(deviation)
```

## Quantitative Comparison Between Predicted and Observed Maximum Horizontal Stress

The **normalized** $\chi^2$ test quantitatively compares the predicted
(`model_shmax()`) and observed $\sigma_{Hmax}$ azimuth relative to the
reported $\sigma$ standard deviation (Wdowinski 1998).

<!-- ```{=tex} -->

<!-- \begin{equation}  -->

<!--   \textrm{Norm} \chi^2_i =  -->

<!--  \frac{\sum_{i=1}^{n} \left(\frac{\alpha_i - \alpha_{predict}}{\sigma_i}\right)^2} -->

<!--  {\sum_{i=1}^{n} \left(\frac{90}{\sigma_i}\right)^2} \end{equation} -->

<!-- ``` -->

<!-- with $\alpha_i$ and $\alpha_{predicted}$ are the observed and predicted -->

<!-- directions of $\sigma_{Hmax}$ with respect to the pole of rotation,  -->

<!-- respectively. -->

<!-- The $\sigma_i$ parameter is the reported uncertainty of the observed -->

<!-- azimuth and $M$ is the number of observations that are used in each -->

<!-- test. -->

The normalized $\chi^2$ test yields a number in the range between 0-1
which indicates the quality of the fit. Low values of the normalized
$\chi^2$ test ($\leq$ 0.15 indicate good agreement between predicted and
observed directions. High values ($>$ 0.7) indicate a systematic misfit
between predicted and observed directions of about 90$^{\circ}$. Random
distribution of $\sigma_{Hmax}$ directions results in Norm
$\chi^2 = 0.33$

The test can be run using `norm_chisq(obs, prd, unc)`. `obs` is a
numeric vector with the observed $\sigma_{Hmax}$; `prd` is a vector with
the predicted $\sigma_{Hmax}$ (vector must be of length of `obs`); and
`unc` is the uncertainty of observed $\sigma_{Hmax}$ (either numeric
vector of length of `obs` or a number).

```{r shmax_test, echo=TRUE}
data("nuvel1") # import example data set for Euler rotations
por <- subset(
  nuvel1, nuvel1$plate.rot == "na"
) # North America relative to Pacific plate
point <- data.frame(lat = 45, lon = 20)

prd <- model_shmax(point, por)
norm_chisq(obs = 90, prd$sc, unc = 10)
```

# Models of current plate motion

The plate motions relative to the Pacific plate according to the
NUVEL-1A model (DeMets et al. 1990) are implemented in the package and
can be imported through:

```{r nuvel1, eval=FALSE, include=TRUE}
data("nuvel1")
head(nuvel1)
```

Other current plate motion models, in particulars NNR-MORVEL-56,
GSRM2.1, REVEL, PB2002, and HS3-NUVEL1A, are implemented and available
through

```{r cpm_models, eval=FALSE, include=TRUE}
data("cpm_models")
head(cpm_models)
```

Any desired relative plate motion can be extracted via the following:

```{r equivalent_rotation, eval=FALSE, include=TRUE}
gsrm <- cpm_models[["GSRM2.1"]]
equivalent_rotation(gsrm, rot = "na", fixed = "eu")
```

# References

DeMets, C., R. G. Gordon, D. F. Argus, and S. Stein. 1990. “Current
Plate Motions” *Geophysical Journal International* 101 (2): 425–78. doi:
[10.1111/j.1365-246x.1990.tb06579.x](https://doi.org/10.1111/j.1365-246x.1990.tb06579.x)

Forsyth, D., and S. Uyeda. 1975. “On the Relative Importance of the
Driving Forces of Plate Motion” *Geophysical Journal International* 43
(1): 163–200. doi:
[10.1111/j.1365-246x.1975.tb00631.x](https://doi.org/10.1111/j.1365-246x.1975.tb00631.x)

Stephan, T., Enkelmann, E., and Kroner, U. (2023). "Analyzing the
horizontal orientation of the crustal stress adjacent to plate
boundaries" *Scientific Reports* (13), 15590.
[doi:[10.1038/s41598-023-42433-2](doi:%5B10.1038/s41598-023-42433-2){.uri}](<https://doi.org/10.1038/s41598-023-42433-2>)

Wdowinski, Shimon. 1998. “A Theory of Intraplate Tectonics” *Journal of
Geophysical Research: Solid Earth* 103 (B3): 5037–59. doi:
10.1029/97jb03390.
<!---[10.1029/97jb03390](https://doi.org/10.1029/97JB03390).-->
